 

### 2021-08-24 vertical-thermal-review.Rmd
schematic 1 updates
Update vertical-thermal-review.Rmd

**Forest responses and feedbacks to climate change will have a critical influence on the future of Earth's climate.**
Globally, temperatures have risen $1.2^\circ$C since 1900 [@rohdeBerkeleyEarthLand2020] and are expected to rise a $1.6 - 4.3^{\circ}$C above pre-industrial levels by 2100 (IPCC 2014), accompanied by increasing severity and frequency of heat waves [@meehlMoreIntenseMore2004] and hotter droughts [@trenberthGlobalWarmingChanges2014b].
These changes are expected to have profound effects on tree metabolism and forest ecosystem function [@breshearsUnderappreciatedPlantVulnerabilities2021; IPBES report], by altering rates of photosynthesis and respiration [@breshearsUnderappreciatedPlantVulnerabilities2021; @corlettImpactsWarmingTropical2011; @scafaroResponsesLeafRespiration2021], causing foliar damage during heat waves [@osullivanThermalLimitsLeaf2017; @corlettImpactsWarmingTropical2011], and reducing growth and elevating mortality during drought [@rollinsonClimateSensitivityUnderstory2020; @breshearsUnderappreciatedPlantVulnerabilities2021; @mcdowellPervasiveShiftsForest2020]. Collectively, these impacts may lead to reduction of forest carbon sequestration, as has already be occurred in some forests [@oishiWarmerTemperaturesReduce2018; @sullivanLongtermThermalSensitivity2020]. The resulting feedbacks to the climate system, carbon storage, and changes in albedo and hydrology will in turn impact the the future trajectory of climate change [@bonan_ecological_2016], yet the degree of the impact remains uncertain [@krauseLargeUncertaintyCarbon2018; @friedlingsteinClimateCarbonCycle2006].

**A great measure of uncertainity arises due to the differential impacts of rising temperatures on trees of differential size and canopy position, especially as because small understory trees exist in microenvironments that are substantially buffered by more exposed canopy trees [@zellweger_seasonal_2019; @davisMicroclimaticBufferingForests2019].**
Forests are vertically and horizontally stratified, containg trees of different sizes along various successional stages. 
Overstory canopies that play a crucial role in moderating forest climatic conditions [@ozanneBiodiversityMeetsAtmosphere2003; @nakamuraForestsTheirCanopies2017], including buffering understory microclimates from extreme temperatures, solar insolation, wind, and other meteorological conditions [@zellweger_seasonal_2019]. This creates a vertical stratification of the biophysical environment such as temperature, light, wind, humidity and CO~2~ concentrations that influences leaf traits, thermoregulation and metabolism along the gradient, with implications for whole plant performance [@fauset_differences_2018; @michaletzEnergeticCarbonEconomic2016].
Despite the fact that this vertical gradient inevitably shapes nearly every aspect of plant metabolism, demography, and ecology, we lack comprehensive understanding of these gradients [but see @niinemetsPhotosyntheticAcclimationSimultaneous2004b].
Importantly, this limits our ability to understand how warming temperatures will affect leaf-level metabolism, whole-plant performance, and, in turn, forest ecosystem dynamics, biodiversity, energy balance, ecosystem function, and biosphere-atmosphere interactions.

@@

![**Figure `r fig_schematic`. Schematic summarizing vertical gradients in forests with dense canopies during daytime growing season conditions.** Patterns summarized here tend to be weaker, or sometimes reversed, under more open canopies, when canopy trees are seasonally deciduous, or at nighttime. Arrows indicate direction of increase, with double-pointed arrows indicating that observations have shown increases both higher and lower in the canopy  [Issue #27](https://github.com/EcoClimLab/vertical-thermal-review/issues/27). ](schematics/Screen Shot 2021-07-19 at 1.10.17 PM.png)

@@

![**Figure `r fig_NEON_vertical`. Vertical gradients in the biophysical environment for six sites six US forest sites with contrasting forest structures in the National Ecological Observatory Network (NEON)**. Height profiles from the growing season are shown for: (a) leaf area density, (b) proportion of sun leaves, (c) proportion light incident at the top of the canopy, and July mean ± 1 standard deviation for (d) maximum photosynthetically active radiation (PAR), (e) maximum wind speed, (f) minimum humidity, (g) maximum $T_{air}$, and (h) maximum biological temperature, $T_{bio}$. Sites include a tropical montane broadleaf evergreen forest (Pu‘u Maka‘ala Natural Area Reserve, Hawai'i; PUUM), a subtropical longleaf pine savanna (Ordway-Swisher Biological Station, FL; OSBS), two temperate broadleaf forests (Smithsonian Conservation Biology Institute, VA, SCBI; Smithsonian Environmental Research Center, MD, SERC), and a mixed northern hardwood and coniferous forest (Harvard Forest, MA; HARV). Vertical profiles in micrometeorological variables (d-h) at all NEON sites are shown in Fig. S1.](NEON_height_profiles/figures/profile_all.png)

**Canopy foliage acts as the primary physical barrier between the atmosphere and the forest floor, buffering multiple aspects of the understory conditions. It is critical in influencing -- and is influenced by -- the vertical biophysical gradient (Fig. `r fig_NEON_vertical`).**
Foliage distribution is determined by the distribution of tree sizes and tree architecture within a forest, and leaf area density (LAD; *i.e.*, leaf area per unit volume) patterns along the vertical gradient are heterogeneous across forests (Figs. `r fig_NEON_vertical`a-b, `r SIfig_NEON_all`). 
Dense canopy forests composed of broadleaf trees can have high LAD in the upper canopy layers, as well as in the understory (e.g., SCBI, SERC and HARV, Fig. `r fig_NEON_vertical`a) -- a pattern also common in tropical forests [@ashtonComparisonsStructureMixed1992; @koikeCanopyStructureTropical1993].
In temperate forests, although LAD is generally more concentrated in the upper-canopy, undualting patterns are formed by conifers and top- vs. bottom-heavy foliage trees, and sometimes by dense understories [e.g., SCBI, Fig. `r fig_NEON_vertical`a, @parkerVerticalProfileCanopy1989; @terborghVerticalComponentPlant1985; @ishiiExploringRelationshipsCanopy2004a; @parkerCanopyLightTransmittance2002]. 
In open canopy forests, including many needle-leaf forests, LAD can be predominantly accumulated in the lower canopy layers [e.g., PUUM and WREF, Fig. `r fig_NEON_vertical`a, @hanberryOpenForestEcosystems2020a; @lawLeafAreaDistribution2001; @baldocchiSeasonalVariationEnergy1997; @bonan_ecological_2016; @ishiiExploringRelationshipsCanopy2004a]. 
Soil moisture conditions, topography, and gap formations following disturbances all alter foliage patterns and further influence the formation of closed vs. open canopy forests [e.g., @almeidaContrastingFireDamage2016; @starkReframingTropicalSavannization2020; @bonan_ecological_2016; @hanberryOpenForestEcosystems2020a]. 
In addition, seasonally dry and wet conditions, deciduousness, and phenology contribute to temporally shifting LAD patterns [@parkerStructuralPhenologyLeaf2004; @smithSeasonalDroughtrelatedChanges2019; @parkerConsequencesEnvironmentalHeterogeneity2019]. 
In this review, we focus on growing season conditions (peak leaf area index, LAI) unless otherwise noted.

@@

```{r eval=FALSE, echo=FALSE}
# refs in this table need to go in the nocite field below (easiest in alphabetical order)
```
---
nocite: | @kafutiFoliarWoodTraits2020; @vanwittenbergheVariabilityStomatalConductance2012a; @roberts_stomatal_1990; @dangProfilesPhotosyntheticallyActive1997; @marencoLeafTraitPlasticity2017; @ambroseEffectsHeightTreetop2010; @zweifelMiddayStomatalClosure2002; @slot_photosynthetic_2019; @hernandezSimilarTemperatureDependence2020; @urbanInductionPhotosynthesisImportance2007; @carterWithinCanopyExperimentalLeaf2018a; @martin_boundary_1999; @mau_temperate_2018; @kosugiSeasonalFluctuationsTemperature2006; @niinemetsVariabilityLeafMorphology2015; @bachofenLightVPDGradients2020; @hamerlynckPhotosyntheticStomatalResponses1996a; @cobleHowVerticalPatterns2016; @wykaResponsesLeafStructure2012; @rijkersEffectTreeHeight2000a; Ishida et al. 1999; @weerasingheCanopyPositionAffects2014; @scartazzaInvestigatingEuropeanBeech2016; @millerOnlySunlitLeaves2021; @harrisChangesLeafProperties2013; @legnerWithincanopyVariationPhotosynthetic2014; @kitaoCanopyNitrogenDistribution2018; @fauset_differences_2018; @rey-sanchez_spatial_2016; @mullerEvidenceEfficientNonevaporative2021; @curtis_intracanopy_2019; @meirLeafRespirationTwo2001; @turnbullScalingFoliarRespiration2003; @arakiVerticalSeasonalVariations2017; @bolstad_foliar_1999; @kenzoHeightrelatedChangesLeaf2015; @harleyEffectsLightTemperature1996;@xuSeasonalVariationTemperature2006;@athertonSpatialVariationLeaf2017; @sackHydrologyLeavesCoordination2003
---

@@

Warming impacts forests, from leaf metabolism to whole forest structure. Leaf level $T_{opt}$ is closely related ecosystem $T_{opt}$. Temperature limitations to leaf processes, such as VPD and gs-limitation, can modify ecosystem $T_{opt}$ if species do not acclimate, resulting in reduced carbon sequestration [@huangAirTemperatureOptima2019; @wayJustRightTemperature2019; @tanOptimumAirTemperature2017; @bennettThermalOptimaGross2021]. At a metabolism level, leaf photosynthesis and respiration are coupled to photosynthetic thermal optima. Beyond this inflection point, photosynthesis declines whereas respiration continues to increase exponentially, eventually shifting the carbon balance from sink to source, independent of water and irradiance [@duffyHowCloseAre2021]. Exposed canopies are likely most vulnerable, as current growing season temperatures are equal to, or higher than photosynthetic optima of temperature and tropical forests [@slotSituTemperatureResponse2017; @mauTemperateTropicalForest2018a; @huangAirTemperatureOptima2019; @kumarathungeAcclimationAdaptationComponents2019], and leaf temperatures may  approach lethal temperatures (e.g., @osullivanThermalLimitsLeaf2017; @tiwariPhotosyntheticQuantumEfficiency2021]. A buffering effect from the canopies might allow shaded layers to photosynthesize longer in the day [@millerOnlySunlitLeaves2021], especially if RH in the understory does not inhibit transpirational cooling.  Thermal limits are especially narrow in tropical and mid-latitudes forests [@osullivanThermalLimitsLeaf2017]. A daytime temperature threshold of 32 degrees C has been suggested for long term climate sensitivity of global tropical forests, where reduced growth, and increased drought and tree mortality will lead to increased carbon source than sink [@sullivanLongtermThermalSensitivity2020].
In the understory, warming is disproportionately affecting the less thermally-adapted functional types, resulting in associated plant community thermophilization [@zellwegerForestMicroclimateDynamics2020; @duqueThermophilizationAdultJuvenile2015; Greiser et al. 2019]. Canopy disturbance (see next section) along with a trend towards younger forests and again reduced tree growth associated with warming [McDowell et al. 2018; REFS] poses an increasing threat to the biodiversity of microrefugia that are otherwise buffered from extremes [@scheffersMicrohabitatsReduceAnimal2013; Greiser et al. 2019].

@@

Canopies are crucial to microclimatic functioning, as they buffer several aspects of forest microenvironments [Zelleweger et al. 2019]. This buffering is influenced by the hydraulogical-cycle at large [Davis et al. 2019], therefore severe droughts place canopies at risk, particularly as larger trees suffer more during droughts[Bennett et al. 2015]. Further, increasing global temperatures show that canopies of temperate and tropical forests are already functioning close to thermal threshold and beyond photosynthetic optimas [@mau_temperate_2018; @huangAirTemperatureOptima2019]. This can have implications in the tropics, where thermal limits are narrower than in temperate forests [Perez and Feeley, 2018].

Canopy structure affects understory species composition and soil fertility, which has been shown to shift under macroclimatic warming and canopy disturbance [@majasalmiImpactTreeCanopy2020; @bertrandCommentForestMicroclimate2020; @maesPlantFunctionalTrait2020]. Macroclimatic warming will disproportionately affect the less warm-adapted functional types within the forest canopy, even with buffering of climatic extremes, thereby creating holes in the canopy that will take a long time to refill (especially in slow-growing tropical sub-canopy specialists), with associated understory plant thermophilization (death is faster than recruitment and growth) [@duqueThermophilizationAdultJuvenile2015; @zellweger_seasonal_2019; @defrenneMicroclimateModeratesPlant2013; @greiserHidingClimateCharacterizing2019]. This imposes an increasing threat to the biodiversity in microrefugias that are  otherwise buffered from extremes [@scheffersMicrohabitatsReduceAnimal2013].

Feedback loops from loss of big trees through natural or anthropogenic disturbance (e.g. @senfCanopyMortalityHas2018a, 2020;  @mcdowellMechanismsPlantSurvival2008; @bennettLargerTreesSuffer2015; @stovallTreeHeightExplains2019a] would increase canopy roughness, and reductions in canopy cover is associated with reduced canopy structural complexity [@senfMappingForestDisturbance2020]. Leading to altered biophysical gradient with local heating effects [@zellweger_forest_2020; @starkReframingTropicalSavannization2020], non-linear threshold responses, causing dramatic shifts from one forest state to another (e.g., transition from forest to savanna-like vegetation in tropical forest regions through "savannization"), with energy balance impacts and implications for forest-atmosphere interactions [@starkReframingTropicalSavannization2020].

Variation in resistance and resilience among forests may drive different trajectories from the forests in the face of global climate change. For example, African rainforests, with canopy trees that are more resilient to warming and drought than other forests [@bennettResistanceAfricanTropical2021]. Temperature acclimation [@slotGeneralPatternsAcclimation2015; @slotPhotosyntheticPlasticityTropical2021; @wayJustRightTemperature2019; @cunninghamTemperateRainforestTrees2003; @cunninghamTemperateRainforestTrees2003] and recovery from canopy heat stress [@smithEmpiricalEvidenceResilience2020] are other mechanisms that will help forests cope with warming.

*For example, species with smaller leaves and larger $gs$ will be less vulnerable to temperature increase than species with larger leaves [@fauset_differences_2018; @songTreeSurfaceTemperature2020].*

-Isoprenes *providing emitting species with a selective advantage in a warming world* [*REF*].

>>>>

**Forest responses and feedbacks to climate change will have a critical influence on the future of Earth's climate.**
Globally, temperatures have risen $1.2^\circ$C since 1900 [@rohdeBerkeleyEarthLand2020] and are expected to reach or even exceed an average of $1.5^\circ$C in the next 20 years [@ipccClimateChangeWidespread2021], accompanied by increasing severity and frequency of heat waves [@meehlMoreIntenseMore2004; @ipccClimateChangeWidespread2021] and hotter droughts [@trenberthGlobalWarmingChanges2014b].
These changes are expected to have profound effects on tree metabolism and forest ecosystem function [@breshearsUnderappreciatedPlantVulnerabilities2021; IPBES report], by altering rates of photosynthesis and respiration [@breshearsUnderappreciatedPlantVulnerabilities2021; @corlettImpactsWarmingTropical2011; @scafaroResponsesLeafRespiration2021], causing foliar damage during heat waves [@osullivanThermalLimitsLeaf2017; @corlettImpactsWarmingTropical2011], and reducing growth and elevating mortality during drought [@rollinsonClimateSensitivityUnderstory2020; @breshearsUnderappreciatedPlantVulnerabilities2021; @mcdowellPervasiveShiftsForest2020]. Collectively, these impacts may lead to reduction of forest carbon sequestration, as has already occurred in some forests [@oishiWarmerTemperaturesReduce2018; @sullivanLongtermThermalSensitivity2020]. The resulting feedbacks to the climate system, carbon storage, and changes in albedo and hydrology will in turn impact the the future trajectory of climate change [@bonan_ecological_2016], yet the degree of the impact remains uncertain [@krauseLargeUncertaintyCarbon2018; @friedlingsteinClimateCarbonCycle2006].

**A great measure of uncertainity arises due to the differential impacts of rising temperatures on trees of differential size and canopy position, especially as because small understory trees exist in microenvironments that are substantially buffered by more exposed canopy trees [@zellweger_seasonal_2019; @davisMicroclimaticBufferingForests2019].**
Forests are vertically and horizontally stratified, containg trees of different sizes along various successional stages. 
Overstory canopies that play a crucial role in moderating forest climatic conditions [@ozanneBiodiversityMeetsAtmosphere2003; @nakamuraForestsTheirCanopies2017], including buffering understory microclimates from extreme meteorological conditions [@zellweger_seasonal_2019]. This creates a vertical stratification of the biophysical environment such as temperature, light, wind, humidity and CO~2~ concentrations that influences leaf traits, thermoregulation and metabolism along the gradient, with implications for whole plant performance [@fauset_differences_2018; @michaletzEnergeticCarbonEconomic2016].
Despite the fact that this vertical gradient inevitably shapes nearly every aspect of plant metabolism, demography, and ecology, we lack comprehensive understanding of these gradients [but see @niinemetsPhotosyntheticAcclimationSimultaneous2004b].
Importantly, this limits our ability to understand how warming temperatures will affect leaf-level metabolism, whole-plant performance, and, in turn, forest ecosystem dynamics, biodiversity, energy balance, ecosystem function, and biosphere-atmosphere interactions.

@@

![**Figure `r fig_schematic`. Schematic summarizing vertical gradients in forests with dense canopies during daytime growing season conditions.** Patterns summarized here tend to be weaker, or sometimes reversed, under more open canopies, when canopy trees are seasonally deciduous, or at nighttime. Arrows indicate direction of increase, with double-pointed arrows indicating that observations have shown increases both higher and lower in the canopy  [Issue #27](https://github.com/EcoClimLab/vertical-thermal-review/issues/27). ](schematics/Schematic figure 1.png)

@@

![**Figure `r fig_NEON_vertical`. Vertical gradients in the biophysical environment for six sites six US forest sites with contrasting forest structures in the National Ecological Observatory Network (NEON)**. Height profiles from the growing season are shown for: (a) leaf area density, (b) proportion of sun leaves, (c) proportion light incident at the top of the canopy, and July mean ± 1 standard deviation for (d) maximum photosynthetically active radiation (PAR), (e) maximum wind speed, (f) minimum humidity, (g) maximum $T_{air}$, and (h) maximum biological temperature, $T_{bio}$. Sites include a tropical montane broadleaf evergreen forest (Pu‘u Maka‘ala Natural Area Reserve, Hawai'i; PUUM), a subtropical longleaf pine savanna (Ordway-Swisher Biological Station, FL; OSBS), two temperate broadleaf forests (Smithsonian Conservation Biology Institute, VA, SCBI; Smithsonian Environmental Research Center, MD, SERC), and a mixed northern hardwood and coniferous forest (Harvard Forest, MA; HARV). Vertical profiles in micrometeorological variables (d-h) at all six NEON sites are shown in Fig. S1.](NEON_height_profiles/figures/profile_all.png)

**Canopy foliage acts as the primary physical barrier between the atmosphere and the forest floor, buffering multiple aspects of the understory conditions. It is critical in influencing -- and is influenced by -- the vertical biophysical gradient (Fig. `r fig_NEON_vertical`).**
Foliage distribution is determined by the distribution of tree sizes and tree architecture within a forest, and leaf area density (LAD; *i.e.*, leaf area per unit volume) patterns along the vertical gradient are heterogeneous across forests (Figs. `r fig_NEON_vertical`a-b, `r SIfig_NEON_all`). 
Dense canopy forests composed of broadleaf trees can have high LAD in the upper canopy layers, as well as in the understory (e.g., SCBI, SERC and HARV, Fig. `r fig_NEON_vertical`a) -- a pattern also common in tropical forests [@ashtonComparisonsStructureMixed1992; @koikeCanopyStructureTropical1993].
In temperate forests, although LAD is generally more concentrated in the upper-canopy, undualting patterns are formed by conifers and top- vs. bottom-heavy foliage trees, and sometimes by dense understories [e.g., SCBI, Fig. `r fig_NEON_vertical`a, @parkerVerticalProfileCanopy1989; @terborghVerticalComponentPlant1985; @ishiiExploringRelationshipsCanopy2004a; @parkerCanopyLightTransmittance2002]. 
In open-canopied forests, including many needle-leaf forests, LAD can be predominantly accumulated in the lower canopy layers [e.g., PUUM and WREF, Fig. `r fig_NEON_vertical`a, @hanberryOpenForestEcosystems2020a; @lawLeafAreaDistribution2001; @baldocchiSeasonalVariationEnergy1997; @bonan_ecological_2016; @ishiiExploringRelationshipsCanopy2004a]. 
Soil moisture conditions, topography, and gap formations following disturbances all alter foliage patterns and further influence the formation of closed vs. open canopy forests [e.g., @almeidaContrastingFireDamage2016; @starkReframingTropicalSavannization2020; @bonan_ecological_2016; @hanberryOpenForestEcosystems2020a]. 
In addition, seasonally dry and wet conditions, deciduousness, and phenology contribute to temporally shifting LAD patterns [@parkerStructuralPhenologyLeaf2004; @parkerConsequencesEnvironmentalHeterogeneity2019]. 
In this review, we focus on growing season conditions (peak leaf area index, LAI) unless otherwise noted.

@@

```{r eval=FALSE, echo=FALSE}
# refs in this table need to go in the nocite field below (easiest in alphabetical order)
```
---
nocite: | @kafutiFoliarWoodTraits2020; @vanwittenbergheVariabilityStomatalConductance2012a; @roberts_stomatal_1990; @dangProfilesPhotosyntheticallyActive1997; @marencoLeafTraitPlasticity2017; @ambroseEffectsHeightTreetop2010; @zweifelMiddayStomatalClosure2002; @slot_photosynthetic_2019; @hernandezSimilarTemperatureDependence2020; @urbanInductionPhotosynthesisImportance2007; @carterWithinCanopyExperimentalLeaf2018a; @martin_boundary_1999; @mau_temperate_2018; @kosugiSeasonalFluctuationsTemperature2006; @niinemetsVariabilityLeafMorphology2015; @bachofenLightVPDGradients2020; @hamerlynckPhotosyntheticStomatalResponses1996a; @cobleHowVerticalPatterns2016; @wykaResponsesLeafStructure2012; @rijkersEffectTreeHeight2000a; @weerasingheCanopyPositionAffects2014; @scartazzaInvestigatingEuropeanBeech2016; @millerOnlySunlitLeaves2021; @harrisChangesLeafProperties2013; @legnerWithincanopyVariationPhotosynthetic2014; @kitaoCanopyNitrogenDistribution2018; @fauset_differences_2018; @rey-sanchez_spatial_2016; @mullerEvidenceEfficientNonevaporative2021; @curtis_intracanopy_2019; @meirLeafRespirationTwo2001; @turnbullScalingFoliarRespiration2003; @arakiVerticalSeasonalVariations2017; @bolstad_foliar_1999; @kenzoHeightrelatedChangesLeaf2015; @harleyEffectsLightTemperature1996;@xuSeasonalVariationTemperature2006; @athertonSpatialVariationLeaf2017; @sackHydrologyLeavesCoordination2003
---

@@

Warming impacts forests, from leaf metabolism to whole forest structure. Leaf level $T_{opt}$ is closely related ecosystem $T_{opt}$. Temperature limitations to leaf processes, such as VPD and gs-limitation, can modify ecosystem $T_{opt}$ if species do not acclimate, resulting in reduced carbon sequestration [@huangAirTemperatureOptima2019; @wayJustRightTemperature2019; @tanOptimumAirTemperature2017; @bennettThermalOptimaGross2021]. At a metabolism level, leaf photosynthesis and respiration are coupled to photosynthetic thermal optima. Beyond this inflection point, photosynthesis declines whereas respiration continues to increase exponentially, eventually shifting the carbon balance from sink to source, independent of water and irradiance [@duffyHowCloseAre2021]. Exposed canopies are likely most vulnerable, as current growing season temperatures are equal to, or higher than photosynthetic optima of temperature and tropical forests [@slotSituTemperatureResponse2017; @mauTemperateTropicalForest2018a; @huangAirTemperatureOptima2019; @kumarathungeAcclimationAdaptationComponents2019], and leaf temperatures may  approach lethal temperatures [e.g., @osullivanThermalLimitsLeaf2017; @tiwariPhotosyntheticQuantumEfficiency2021]. A buffering effect from the canopies might allow shaded layers to photosynthesize longer in the day [@millerOnlySunlitLeaves2021], especially if RH in the understory does not inhibit transpirational cooling.  Thermal limits are especially narrow in tropical and mid-latitudes forests [@osullivanThermalLimitsLeaf2017]. A daytime temperature threshold of 32 degrees C has been suggested for long term climate sensitivity of global tropical forests, where reduced growth, and increased drought and tree mortality will lead to increased carbon source than sink [@sullivanLongtermThermalSensitivity2020].
In the understory, warming is disproportionately affecting the less thermally-adapted plant species, resulting in associated plant community thermophilization [@zellwegerForestMicroclimateDynamics2020; @duqueThermophilizationAdultJuvenile2015; @greiserHidingClimateCharacterizing2019]. Canopy disturbance (see next section) along with a trend towards younger forests and again reduced tree growth associated with warming [McDowell et al. 2018; REFS] poses an increasing threat to the biodiversity of microrefugia that are otherwise buffered from extremes [@scheffersMicrohabitatsReduceAnimal2013; @greiserHidingClimateCharacterizing2019].

@@

Canopies are crucial to microclimatic stratification, as they buffer several aspects of forest microenvironments [Figs 1-2, @zellweger_seasonal_2019]. This buffering is influenced by hydrology, as wetter sites and seasons allow for denser, and more buffered canopies  [@davisMicroclimaticBufferingForests2019]. Severe droughts and rising temperatures place canopies at risk, as larger trees tend to suffer more during droughts [@bennettLargerTreesSuffer2015]. The disproportionate impact of disturbance on large trees holds for other types of disturbance, including forest fragmentation and logging (Laurance et al. 2006; Miller et al. 2004), although fire tends to predominantly affect smaller trees (Brando et al. 2019). Canopy structure affects understory species composition, which has been shown to shift under warming and canopy disturbance [@majasalmiImpactTreeCanopy2020; @bertrandCommentForestMicroclimate2020; @maesPlantFunctionalTrait2020]

In general, disturbances (e.g., drought, fire, fragmentation, and logging) reduce canopy cover, which in turn increases incident radiation levels and temperatures within the canopy [@starkReframingTropicalSavannization2020]. This shift to hotter and drier microclimates often makes forests more susceptible to further disturbances, for example, increasing fire risk (Brando et al. 2014; Aragao et al. 2018). Severe degradation impacts can cause dramatic ecological state changes (e.g., the transition from forest to more open, savanna-like vegetation in tropical forest regions through 'savannization'; Silverio et al. 2013) and non-linear threshold responses in energy balance and associated microclimates, with implications for forest-atmosphere interactions [@starkReframingTropicalSavannization2020].

Variation in resistance and resilience among forests may drive different trajectories in the face of global climate change. For example, African rainforests, with canopy trees that are more resilient to warming and drought than other forests [@bennettResistanceAfricanTropical2021]. Species with smaller leaves and larger gs will be less vulnerable to temperature increase than species with larger leaves [@fauset_differences_2018; @songTreeSurfaceTemperature2020] and isoprene emitting species will have a selective advantage in a warming world [@taylorCapacityEmitIsoprene2019]. Temperature acclimation [@slotGeneralPatternsAcclimation2015; @slotPhotosyntheticPlasticityTropical2021; @wayJustRightTemperature2019;  @cunninghamTemperateRainforestTrees2003] and recovery from canopy heat stress [@smithEmpiricalEvidenceResilience2020] will also help forests cope with warming.
